\myTitle

### 2024-02-26 body-text-of-paper.tex
Eric edited through the end of the methods.

Figure~\ref{fig:gsp1} shows an example GSP for the simple case
of simulating $F_2$ individuals. The figure and its caption
should be studied, as the following section
uses terminology that is defined in the figure caption.

It is important to understand that, apart from the founder nodes, the individual
nodes in a GSP do not necessarily represent only a single individual.  Specifically,
since we are preserving all genetic material, multiple gametes may get segregated
through each non-founder node in a GSP\@.

@@

In the \gscramble{} package there is a convenience function called {\footnotesize\tt create\_GSP()} that
will create GSPs for samples up to and including second-generation backcrosses (an example appears in Figure~\ref{fig:gsp2});
however, \gscramble{} is designed
to handle user-specified GSPs of arbitrary complexity.  When creating a unique GSP,
the following requirements are necessary for the GSP to be valid
and the simulation to execute successfully.
Writing $\founders$ for the set of all founder nodes, $\nonfound$ for the set of all non-founder nodes, and $\nonfounds$ for all non-founder
nodes that are not parents of other non-founders, but are adjacent to a sample node, $\nonfounde$ for all
non-founder nodes that have non-founder node daughters, but no sample nodes, and $\nonfoundse$ for all
non-founder nodes that are parents of non-founder nodes and also adjacent to sample-nodes, the necessary conditions
for a valid GSP, in addition to having no inbreeding loops, are:
\begin{equation}
\begin{aligned}
\sum_{x \in \nonfounds \cup \nonfoundse} S_x &= |\founders| &  \\
\sum_{i=1}^{e_x} G^-_{x,i} &= 2 & \forall x \in \founders \\
G^+_{x,1} &= G^+_{x,2} & \forall x \in \nonfound \\
\sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & \forall x \in \nonfounde  \\
 S_x/2 + \sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & ~~~\forall x \in \nonfoundse.  \\
\end{aligned}
\end{equation}
Before simulating from a GSP, \gscramble{} checks to ensure the GSP
satisfies these validity conditions.  The simulation of genomic segments
from founders to samples in a GSP is performed by the \gscramble{} function
{\footnotesize\tt segregate()}, which requires as input a list of GSPs through which
to segregate genomic segments, and, for each such GSP, a table that maps the
population labels on each GSP (in the inverted triangles atop the founders) to the
different sampled groups of individuals.

@@

We illustrate the use of \gscramble{} in three example cases. The first is
a simple simulatino to show that, unlike sampling with replacement, the
approach of \gscramble{} does not produce RISPI.
The next two are examples using empirical data sets.
The first of these involves
simulating a small number ($<\,$70) of markers that are fixed, or nearly
fixed, for alternate alleles between two species of trout: rainbow trout ({\em 
Oncorhynchus mykiss}) and cutthroat trout ({\em O. clarkii}). The goal of the 
simulation study is to determine how accurately individuals of different hybrid
categories between the species can be identified, and whether the use of linked 
markers deflates the estimates of uncertainty to an unacceptable degree.  The 
second example involves many more markers ($\approx$14,000) and 
investigates the accuracy for
identifying individuals admixed between closely related groups of feral
pigs in Missouri.

@@

For this example, we repeat the simulation described in \nameref{sec:rispi-sim};
however, after simulating individuals in the two samples drawn from the same population,
we create data sets of admixed individuals using \gscramble{}, rather than sampling
gene copies with replacement.  As before, individuals were simulated with fractions of
ancestry from sample $A$ in $\{0, \frac{1}{8}, \frac{1}{4}, \frac{3}{8}, \frac{1}{2}, \frac{5}{8}, \frac{3}{4}, \frac{7}{8}, 1\}$, but these were simulated from pedigrees using gscramble.  We limited
this simulation to one of the scenarios that showed extreme RISPI when sampling gene
copies with replacement: $L=1,000$ markers with sizes of samples $A$ and $B$ being
$N=50$. This simulation was repeated 240 times, and the results of each analyzed with
ADMIXTURE to estimate the ancestry fractions of each simulated individual.

@@

\citeauthor{rizza2023limited} mapped the 65 markers to the Omyk\_v1.0 genome assembly (RefSeq 
Accession: GCF\_002163495.1), so we used
the linkage map described in \citep{pearse2019sex} that was prepared by linkage mapping 
of $\approx$57,000 markers described in \citet{palti2015development} to Omyk\_v1.0. We mapped the $\approx$57,000 markers to Omyk\_v1.0, and then resolved inconsistencies to produce a map
of 32,923 markers throughout the genome and rendered it to plink format for use with \gscramble{}.

In the original paper, \citet{rizza2023limited} identified 15 $F_1$s,
1 $F_2$,
6 cutthroat backcrosses ($BC1$s) and 
6 second-generation cutthroat backcrosses ($BC2$s),
and 1 fish of an unknown hybrid category that may have been the
product of a $BC2$ crossed with a $BC1$.   In our simulations, we sought to simulate
a similar collection of hybrid individuals, so, for each replicate, we used
\gscramble{} to simulate 16 $F_1$s, 4 $F_2$s, 6 $BC1$s, 4 $BC2$s, 4 $BC3$s,  4 individuals
that were the products of a $BC1$ mating with a $BC2$, and finally, 4 individuals that were simulated pure cutthroat. Individuals not used as founders
in \gscramble{} were retained as pure cutthroat and pure steelhead.  The goal of the simulation
was to assess how well these different hybrid categories can be distinguished from one another
when using either all 65 markers, or just the 30 markers occurring on different chromosome arms
used by  \citet{rizza2023limited}.

@@

The 160 invasive wild pig samples represented 120 reference individuals from the core regions of
3 populations (40 individuals each informed by previous analyses)
and 40 individuals from the contact regions of these populations.
We used ADMIXTURE version 1.3.0 \citep{alexander2009fast} to characterize the genetic structure
of the 160 samples (for $K = {1...6}$). Using the 120 reference individuals identified to represent population cores,
we sought to classify the admixture patterns of the 40 empirical samples
from the contact regions using simulated pedigrees from \gscramble{}. To address this objective we simulated
pedigrees to represent the likely classes of admixed individuals that would be present in the samples
from the contact regions of the 3 reference populations. The types of admixed individuals we considered were
$F_1$, $F_2$, $\BC_1$ and $\BC_2$ representing all two-population combinations
(Pop1 x Pop2, Pop1 x Pop3, Pop2 x Pop3, and reciprocally Pop2 x Pop1, Pop3 x Pop1, Pop3 x Pop2 as required
to simulate backcrossed classes).
We also simulated individuals from each of the founder populations. The first step of this workflow involved
simulating admixed individuals for 2 founder populations using one of the preset \gscramble{}
GSPs provided by the function {\footnotesize\tt create\_GSP()} for producing each of the types of admixed
individuals. Although we could have simulated all 4 types of admixed individuals from a single GSP (Figure~\ref{fig:gsp2}),
different numbers of simulated individuals are produced for each hybrid class. Therefore, for the sake of the
computational efficiency from simulating the exact number of desired individuals, we chose to simulate from
a single GSP for each of the hybrid classes by setting the corresponding hybrid parameter in {\footnotesize\tt create\_GSP()} to {\footnotesize\tt TRUE}. We used the default settings of the {\footnotesize\tt plink\_map2rec\_rates()} to specify the recombination rates. The 120 simulated genotypes were written to PLINK .ped and .map files using {\footnotesize\tt gscramble2plink()}. We used PLINK version 1.9 \citep{purcell2007plink} to convert the genotypes to binary files and used these as input into ADMIXTURE for which we specified $K=3$ to produce ancestry values. We iteratively
performed this workflow to produce 1,000 simulated genotypes and ancestry values for each of the
4 hybrid classes, as well as founder individuals, for each of the population combinations, for a total of
21,000 simulated individuals.

>>>>

Figure~\ref{fig:gsp1} shows an example GSP for the simple case
of simulating $F_2$ individuals. The figure and its caption
should be studied, as the following section
uses terminology that is defined in the figure caption.
It is important to understand that, apart from the founder nodes, the individual
nodes in a GSP do not necessarily represent only a single individual.  Specifically,
since we are preserving all genetic material, multiple gametes may get segregated
through each non-founder node in a GSP\@.

@@

In the \gscramble{} package there is a convenience function called {\footnotesize\tt create\_GSP()} that
will create GSPs for samples up to and including second-generation backcrosses (an example appears in Figure~\ref{fig:gsp2});
however, \gscramble{} is designed
to handle user-specified GSPs of arbitrary complexity.  When creating a unique GSP,
there are several requirements to ensure that the GSP is valid
and the simulation can execute successfully.
Writing $\founders$ for the set of all founder nodes, $\nonfound$ for the set of all non-founder nodes, and $\nonfounds$ for all non-founder
nodes that are not parents of other non-founders, but are adjacent to a sample node, $\nonfounde$ for all
non-founder nodes that have non-founder node daughters, but no sample nodes, and $\nonfoundse$ for all
non-founder nodes that are parents of non-founder nodes and also adjacent to sample-nodes, the necessary conditions
for a valid GSP, in addition to having no inbreeding loops, are:
\begin{equation}
\begin{aligned}
\sum_{x \in \nonfounds \cup \nonfoundse} S_x &= |\founders| &  \\
\sum_{i=1}^{e_x} G^-_{x,i} &= 2 & \forall x \in \founders \\
G^+_{x,1} &= G^+_{x,2} & \forall x \in \nonfound \\
\sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & \forall x \in \nonfounde  \\
 S_x/2  &= G^+_{x,1} +  G^+_{x,2} & ~~~\forall x \in \nonfounds \\
 S_x/2 + \sum_{i=1}^{e_x} G^-_{x,i} &= G^+_{x,1} +  G^+_{x,2} & ~~~\forall x \in \nonfoundse.  \\
\end{aligned}
\end{equation}
In words, the separate lines above specify that 1) the number of samples
must equal the number of founders; 2) exactly two gametes must be
segregated from each founder; 3) the number of gametes incoming on each of the
two edges coming into a non-founder (from its parents) must be equal; 
4) the number of gametes going downward, out of a non-founder with no sample
daughters, must equal the number of gametes coming into that non-founder;
5)  the total number of gametes entering a non-founder with only sample daughters
must be half the number of sample daughters; and 6) the number of gametes
entering non-founders with both sample daughters and non-founder daughters must
equal half the number of its sample daughters plus the sum of the number of gametes segregated
along each edge to all of its non-founder daughters.

Before simulating from a GSP, \gscramble{} checks to ensure the GSP
satisfies these validity conditions.  The simulation of genomic segments
from founders to samples in a GSP is performed by the \gscramble{} function
{\footnotesize\tt segregate()}, which requires as input a list of GSPs through which
to segregate genomic segments, and, for each such GSP, a table that maps the
population labels on each GSP (in the inverted triangles atop the founders) to the
different sampled groups of individuals.

@@

We illustrate the use of \gscramble{} in three example cases. The first is
a simple simulation to show that, unlike sampling with replacement, the
approach of \gscramble{} does not produce RISPI.
The next two are examples using empirical data sets.
The first of these involves
simulating a small number ($<\,$70) of markers that are fixed, or nearly
fixed, for alternate alleles between two species of trout: rainbow trout ({\em 
Oncorhynchus mykiss}) and cutthroat trout ({\em O. clarkii}). The goal of the 
simulation study is to determine how accurately individuals of different hybrid
categories between the species can be identified, and whether the use of linked 
markers deflates the estimates of uncertainty to an unacceptable degree.  The 
second example involves many more markers ($\approx$14,000) and 
investigates the accuracy for
identifying individuals admixed between closely related groups of feral
pigs in Missouri.

@@

For this example, we repeat the simulation described in {\em \nameref{sec:rispi-sim}};
however, after simulating individuals in the two samples drawn from the same population,
we create data sets of admixed individuals using \gscramble{}, rather than sampling
gene copies with replacement.  As before, individuals were simulated with fractions of
ancestry from sample $A$ in $\{0, \frac{1}{8}, \frac{1}{4}, \frac{3}{8}, \frac{1}{2}, \frac{5}{8}, \frac{3}{4}, \frac{7}{8}, 1\}$, but these were simulated from pedigrees using gscramble.  We limited
this simulation to one of the scenarios that showed extreme RISPI when sampling gene
copies with replacement: $L=1,000$ markers with sizes of samples $A$ and $B$ being
$N=50$. This simulation was repeated 240 times, and the results of each replicate were analyzed with
ADMIXTURE to estimate the ancestry fractions of each simulated individual.

@@

\citeauthor{rizza2023limited} mapped the 65 markers to the Omyk\_v1.0 genome assembly (RefSeq 
Accession: GCF\_002163495.1), so we used
the linkage map described in \citep{pearse2019sex} that was prepared by linkage mapping 
of $\approx$57,000 markers described in \citet{palti2015development} to Omyk\_v1.0. We mapped these $\approx$57,000 markers to Omyk\_v1.0, and then resolved inconsistencies to produce a map
of 32,923 markers throughout the genome and rendered it to plink format for use with \gscramble{}.

In the original paper, \citet{rizza2023limited} identified 15 $F_1$s,
1 $F_2$,
6 cutthroat backcrosses ($BC_1$s) and 
6 second-generation cutthroat backcrosses ($BC_2$s),
and 1 fish of an unknown hybrid category that may have been the
product of a $BC_2$ crossed with a $BC_1$.   In our simulations, we sought to simulate
a similar collection of hybrid individuals, so, for each replicate, we used
\gscramble{} to simulate 16 $F_1$s, 4 $F_2$s, 6 $BC_1$s, 4 $BC_2$s, 4 $BC_3$s,  4 individuals
that were the products of a $BC_1$ mating with a $BC_2$, and finally, 4 individuals that were simulated pure cutthroat. Individuals not used as founders
in \gscramble{} were retained as pure cutthroat and pure steelhead.  The goal of the simulation
was to assess how well these different hybrid categories can be distinguished from one another
when using either all 65 markers, or just the 30 markers occurring on different chromosome arms
used by  \citet{rizza2023limited}.

@@

The 160 invasive wild pig samples represented 120 reference individuals from the core regions of
3 populations (40 individuals each informed by previous analyses)
and 40 individuals from the contact regions of these populations.
We used ADMIXTURE version 1.3.0 \citep{alexander2009fast} to characterize the genetic structure
of the 160 samples (for $K = {1...6}$). Using the 120 reference individuals identified to represent population cores,
we sought to classify the admixture patterns of the 40 empirical samples
from the contact regions using simulated pedigrees from \gscramble{}. To address this objective we simulated
pedigrees to represent the likely classes of admixed individuals that would be present in the samples
from the contact regions of the 3 reference populations. The types of admixed individuals we considered were
$F_1$, $F_2$, $\BC_1$ and $\BC_2$ representing all two-population combinations
(Pop1 x Pop2, Pop1 x Pop3, Pop2 x Pop3, and reciprocally Pop2 x Pop1, Pop3 x Pop1, Pop3 x Pop2).
We also simulated individuals from each of the founder populations. The first step of this workflow involved
simulating admixed individuals for 2 founder populations using one of the preset \gscramble{}
GSPs provided by the function {\footnotesize\tt create\_GSP()} for producing each of the types of admixed
individuals. Although we could have simulated all 4 types of admixed individuals from a single GSP (Figure~\ref{fig:gsp2}),
different numbers of simulated individuals are produced for each hybrid class. Therefore, for the sake of the
computational efficiency from simulating the exact number of desired individuals, we chose to simulate from
a single GSP for each of the hybrid classes by setting the corresponding hybrid parameter in {\footnotesize\tt create\_GSP()} to {\footnotesize\tt TRUE}. We used the default settings of the {\footnotesize\tt plink\_map2rec\_rates()} to specify the recombination rates. The 120 simulated genotypes were written to PLINK .ped and .map files using {\footnotesize\tt gscramble2plink()}. We used PLINK version 1.9 \citep{purcell2007plink} to convert the genotypes to binary files and used these as input into ADMIXTURE for which we specified $K=3$ to produce ancestry values. We iteratively
performed this workflow to produce 1,000 simulated genotypes and ancestry values for each of the
4 hybrid classes, as well as founder individuals, for each of the population combinations, for a total of
21,000 simulated individuals.
